Article Title

### 2013-12-21 plos.tex
Rev28 -- Kreso's edits of Abstract--Results.

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser subsets of neurons \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of the functional organization of neural circuits.  However, estimating and interpreting large correlation matrices obtained from finite recordings is challenging.  Estimation can be improved by regularization: biasing of the usual unbiased estimate toward its low-dimensional approximation.  The amount of improvement is greatest when the approximation chosen as the target of regularization parsimoniously captures the dominant dependencies in the data.  Therefore, the construction of the most efficient estimator of neural correlation matrices is an empirical question that informs about the types of dominant dependencies governing the system.

In this study, we sought most statistically efficient estimators of neural correlations in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes of about 100 $\mu$m in diameter (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix would be most efficiently represented by a sparse network of partial correlations between pairs of observed neurons combined with common latent factors representing unobserved inputs and global network fluctuations.  Indeed, in cross-validation tests, the covariance matrix estimator based on such approximations outperformed other regularized estimators. Mirroring previously described patterns of synaptic connectivity, the density of positive partial correlations decreased rapidly with both physical distances and preferred orientation differences whereas negative partial correlations were less selective.  The inferred function connectivity obtained using these methods can provide hypotheses about synaptic architecture that can guide future experimental work.
\section*{Author Summary}
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
Correlations in the activity of neurons have proven useful as descriptors of the functional organization of neural circuits with implications for stimulus coding and circuit architecture.  Estimation of correlation matrices can be improved by imposing some structure on the estimate. Greatest improvements are attained when the imposed structure efficiently captures real dependencies in the data. Using fast 3D two-photon imaging of calcium signals, we recorded the activity of large and dense groups of cells in mouse visual cortex and evaluated the performance of correlation matrix estimators that imposed different kinds of structure. The correlation structure of the estimator that led to greatest improvement comprised a sparse network of partial correlations between pairs of neurons combined with partial correlations with several common latent factors. Both components were necessary for efficient estimation. Since it leads to better estimates, we proposed that this sparse+latent structure provides a better picture of the  functional connectivity than that obtained from the usual sample correlation matrix in densely sampled neural recordings. As an application of this approach, we analyzed how the inferred connectivity related to distances between cells and differences in their preferred orientations and found basic agreement with previous studies of synaptic connectivity.

@@

Alternatively, the correlation matrix can be formulated through the partial correlations between pairs of neurons. The partial correlation between a pair of neurons is the linear correlation remaining after accounting for correlations to all the other neurons. Partial correlations carry particular significance when conditional dependencies between all the variables are thought to be dominated by linear effects. In such systems, zero  partial correlation between two variables implies their conditional independence.  Conditional independence between a pair of variables suggests a lack of direct interaction between the underlying processes. Therefore, estimation of networks of partial correlations (sometimes called \emph{association networks}) have been used to infer interactions between components in complex systems such as gene interaction networks \cite{Schafer:2005,Peng:2009} and functional connectivity between brain regions in fMRI signals \cite{Varoquaux:2012,Ryali:2012}. In \emph{elliptical distributions} (those of the form $x \sim \frac 1 Z g\left((x-\mu)^\T\Sigma^{-1}(x-\mu)\right)$, where $g$ is an arbitrary function, $Z$ is the scalar normalizer, $\Sigma$ is the covariance matrix) and the multivariate normal distribution in particular, interactions are completely defined by linear effects and partial interactions express conditional dependencies. A multivariate normal distribution specified through its graph of non-zero partial correlations is known as a Gaussian Graphical Model or Gauss-Markov Random Field \cite{Koller:2009}. With increasing departure from linearity, the correspondence between conditional dependence and partial correlations diminishes or completely breaks down \cite{Loh:2012}. For example, the coupling terms in pairwise Ising models have no direct relationship to the partial correlations \cite{Schneidman:2006,Tkacik:2006}.

\subsection*{Estimation of neural correlation matrices}
In this study, we pursued two related aims: (a)~improved estimation of neural correlation matrices and (b)~discovery of low-dimensional structure of correlations in recordings of multineuronal activity to facilitate their interpretation.

The true correlation matrix is the normalized version of the true covariance matrix defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; $x$ is the $p\times 1$ vector of real-valued instantaneous firing rates in bins of duration $\Delta t$; and $\mu$ is the vector of mean firing rates. In the case of noise correlations, $\mu$ depends on the stimulus condition.  The usual estimator of the covariance matrix is the \emph{sample covariance matrix} $C_{\sf 0}$ computed from the empirical sample of observations $x(t),\; t=n(1,\ldots,n)$ as
\begin{equation}
    C_{\sf 0} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ is the number of degrees of freedom per neuron in the sample ($\nu=n-1$ if observations are independent). In this study, we keep the sample mean $\bar x$ as the estimate of the true mean $\mu$, but seek a better estimate of the covariance matrix than $C_{\sf 0}$.

@@

Estimation of the covariance matrix from a large population presents a number of numerical challenges.  The amount of recorded data grows only linearly with the population size whereas the number of estimated coefficients increases quadratically.  This mismatch leads to increased opportunities for spurious correlations, to overestimation of shared activity (\emph{i.e.}\;overestimation of large eigenvalues) \cite{Ledoit:2004}, and to poorly conditioned estimates of the partial correlations \cite{Schafer:2005}.

Estimation can be improved through \emph{regularization}: the deliberate biasing of the estimate toward a low-dimensional approximation \cite{Schafer:2005,Bickel:2006}.  The usual estimate $C_{\sf 0}$ has the advantage of being unbiased ($\E{C_0}=\Sigma$) but, on average, it falls far from $\Sigma$ due to its sensitivity to sampling noise in training sample.  Low-dimensional estimates of various forms are typically less susceptible to sampling noise but are also liable to introduce their respective biases away from truth $\Sigma$.  Regularization works by striking a favorable balance between bias and variability. Regularization can produce \emph{some} improvement even with an arbitrary target estimate that has no relation to the true dependencies in the data.  Yet, when the target estimate is well suited for capturing the important features of the true covariance matrix with few terms, it will introduce minimal bias and outperform other estimators.

\subsection*{Illustration of estimation of a neural correlation matrix}
To illustrate challenges of estimating the correlation matrix from a finite sample, we briefly consider a regularization scheme based on \emph{covariance selection} (Figure \ref{fig:01}) \cite{Dempster:1972}. In covariance selection, the estimate is produced by fitting only an optimal subset of the coefficients of the precision matrix while setting the rest to zero. We estimated the covariance matrix of a period of a somatic calcium signals of a group of 298 neurons in mouse visual cortex using both the sample covariance matrix (Fig.\;\ref{fig:01}\,A, B) and covariance selection (Fig.\;\ref{fig:01}\,C). Due to the low-pass filtration effect of calcium dye kinetics, correlations in unprocessed calcium signals are higher than typical firing rate correlations on shorter temporal scales. The close similarity of the two estimates (Fig.\;\ref{fig:01}\,B and C) belies the utterly different partial correlation structure (Fig.\;\ref{fig:01}\,D, E): the regularized estimate is produced from the same data by zeroing the optimal set of 31,501 (71.2\%) of the off-diagonal coefficients of the precision matrix and fitting the remaining 12,752 coefficients. Estimation by covariance selection is attractive for several reasons: First, it has fewer free parameters and is therefore less susceptible to sampling noise, yielding estimates that are, on average, closer to truth $\Sigma$. Additionally, if we assume that partial correlations do indeed reflect elements of the functional connectivity in the circuit, the second estimate also appears more plausible: in neural circuits, each neuron only interacts with a fraction of the entire population.

@@

\subsection*{Model selection}
Covariance selection is but one of many conceivable low-dimensional approximations of the covariance matrix. Indeed any probabilistic model of the joint population activity regularized toward a small number of parameters automatically defines a low-dimensional approximation of the covariance matrix. Models that accurately reflect the dominant functional dependencies in the circuit, its so-called `functional connectivity', will tend to produce estimates that are nearest to the true value of the covariance matrix and are said to be more \emph{efficient} than the other estimators or to \emph{dominate} the other estimators. An estimator's superior efficiency can be ascertained by cross-validation.  Because different types of neural circuits are governed by different types of dependencies, different probabilistic models and covariance matrix estimators may dominate in each case. The estimator that is shown to dominate all others for a specific type of neural circuit may be proposed as the best descriptor of the low-dimensional representation of the correlation structure of the system of interest. Its structure may then be analyzed and related to the circuit's anatomical organization.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Estimators whose low-dimensional regularization targets can represent the structure of the true covariance matrix outperform other estimators.}
        {\bf Row 1.} Graphical representations of four types of low-dimensional structures of interactions between observed neurons (green spheres) and latent units (light-shaded spheres).
        In the `independent model' ({\bf A}), observed neurons exert no linear effects on one another neither directly nor through interactions with common latent units. 
        In `latent factors' ({\bf B}), the correlated activity of observed cells is driven by several latent units. 
        In `sparse interactions' ({\bf C}), the correlation matrix is defined by a set of linear interactions between observed neurons. 
        In `sparse+latent' ({\bf D}), correlations arise through direct linear interactions between some pairs of observed neurons and through interactions with common latent units. 
        {\bf Row 2.} Examples of $50\times 50$ correlation matrices corresponding to each type of low-dimensional structure. 
        The factor model ({\bf B}) has three latent units. 
        The partial correlation matrix of the sparse model ({\bf C}) is 73\% sparse.
        The `sparse+latent' matrix has one latent unit and its direct interactions are 78\% sparse.
        {\bf Row 3.} Examples of sample correlation matrices calculated from samples of 1000 observations taken from simulated random processes with corresponding correlation matrices from row 2.
        {\bf Row 4.} Mean \emph{excess loss} (Eq.~\ref{eq:excess-loss}) attained by each of the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.
        {\bf Row 5.} Mean \emph{cross-validation loss} (Eq.~\ref{eq:rel-cv-loss}) of covariance estimators with respect to the matching estimator . The values are relative to the validation loss of the estimator that matches the low-dimensional structure of the true covariance matrix. The error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

In the first regularized estimator $C_{\sf diag}$, the target estimate is the diagonal matrix $D$ containing on its diagonal estimates of the variances. The regularized estimate is obtained by linear \emph{shrinkage} of the unbiased estimate $C_{\sf 0}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}
C_{\sf diag} = (1-\lambda) C_{\sf 0} + \lambda D
\end{equation}
The diagonal target estimate expresses the idea of a lack of dependence (or of linear association) between the activity of observed neurons (Fig.~\ref{fig:02}\,A).  
If this assumption aptly describes recorded data, then strong shrinkage toward $D$ will add little bias while strongly reducing the variability of the estimate. Shrinkage allows for partial commitment to the low-dimensional representation.

In the second regularized estimator $C_{\sf factor}$, the target estimate is the factor model $F =  L L^\T + \Psi$ with $d$ factors so that $L$ is the $p\times d$ matrix of \emph{factor loadings} and the diagonal matrix $\Psi$ contains the independent variances of each neuron.
Then the estimate is 
\begin{equation}
C_{\sf factor} = (1-\lambda) C_{\sf 0} + \lambda F
\end{equation}
This estimator has two hyperparameters: the number of factors $d$ and shrinkage intensity $\lambda$. The target estimate $F$ expresses the assumption that correlated fluctuations in the population activity are driven by a small number of latent factors that affect many cells while direct interactions between cells are insignificant (Fig.~\ref{fig:02}\,B).

The third estimator $C_{\sf sparse}$ is based on the assumption that all correlations are the result of direct linear interactions between pairs of observed cells and that such interactions occur only between a fraction of such pairs (Fig.~\ref{fig:02}\,C).  This assumption is enforced by setting to zero an optimal subset of the off-diagonal elements of the precision matrix.  Therefore, $C_{\sf sparse}$ is biased toward the assumption that correlations arise due to linear interactions between pairs of recorded neurons (Fig.~\ref{fig:02}\,C). Then the estimator is 
\begin{equation}
C_{\sf sparse} = S^{-1}
\end{equation}
where $S$ is a sparse matrix with a large fraction of zeros in its off-diagonal elements. The estimate has one hyperparameter to regulate the sparsity (fraction of off-diagonal zeros) in $S$.

Finally, we consider the fourth estimator $C_{\sf sparse+latent}$, which provides for both common latent factors interacting with all recorded neurons and sparse interactions between the recorded neurons (Fig.~\ref{fig:02}\,D). This estimator has the form \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}
C_{\sf sparse+latent} = (S - LL^\T)^{-1}
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a $d\times p$ matrix of factor loadings. The estimator has two hyperparameters: the number of latent units $d$ and the sparsity of $S$.

\subsection*{Simulation}
To verify our approach and to illustrate the performance of the four regularized estimators, we constructed four model populations of size $p=50$~neurons each with correlation structures of the same type as the target estimates of the first (Fig.~\ref{fig:02} Row 2). Panel A2 contains a diagonal correlation matrix matching the target of $C_{\sf diag}$. Panel B2 is a factor model with 3 factors conforming to the family of target estimates of $C_{\sf factor}$. Panel B3 is a correlation matrix with sparse partial correlations (76\% off diagonal zeros in the precision matrix). Finally, Panel B4 is a correlation matrix whose inverse is composed the sum of a sparse matrix (82\% sparse) and a low-rank component (rank=1).
Row 3 contains sample correlation matrices calculated from samples of size $n=1000$ from a multivariate normal distributions with the respective correlation matrices from Row 2.

To evaluate the performance of a covariance matrix $C$, we define a real-valued \emph{loss function} $\loss{C,\Sigma}$ such that it attains its minimum when $C=\Sigma$.  The loss function quantifies the error of the estimate, \emph{i.e.}~the deviation of the estimate $C$ from truth $\Sigma$.

@@

This choice is motivated by mathematical convenience. Other popular choices for the loss function are the Frobenius norm of the difference $\|C-\Sigma\|_F$ \cite{Ledoit:2004,Schafer:2005}, Stein's entropy loss, and quadratic loss \cite{James:1961,Fan:2008}.  We expect that the main conclusions of our study will not change qualitatively under other well behaved loss functions.

To make the loss function to assume zero at its minimum when $C=\Sigma$, we define \emph{excess loss} as
\begin{equation}\label{eq:excess-loss}
    \eloss{C,\Sigma} = \loss{C,\Sigma}-\loss{\Sigma,\Sigma}
\end{equation}

In Figure \ref{fig:02}, row 4 shows the mean excess losses and their standard errors calculated from 30 samples of sizes n=250, 500, 1000, 2000, and 4000 for each estimator and each kind of ground truth. For each estimator, its hyperparameters were estimated by nested cross-validation (See Methods for more details.)

As expected, estimators with matching low-dimensional structures typically outperformed the other estimators. There were two exceptions to this observation. For small sample sizes, before the data were sufficient to reveal the true correlation structure and to  allow the correct model to dominate, estimates with simpler targets often outperformed the matching estimator.

When ground truth $\Sigma$ is not accessible, loss can be estimated solely from the data through \emph{validation}.  In validation, an additional independent \emph{testing sample} is used to compute an independent sample covariance estimate $C_{\sf 0}^\prime$.  Then \emph{validation loss} $\loss{C,C_{\sf 0}^\prime}$ can be used as an unbiased estimate\footnote
{
    Validation loss $\loss{C,C_{\sf 0}^\prime}$ is an unbiased estimate of loss $\loss{C,\Sigma}$ when $\loss{\cdot,\cdot}$ is \emph{additive} in its second argument so that 
 \begin{equation*}\label{eq:additivity}
 \loss{C,z_1} + \loss{C,z_2} \equiv \loss{C,z_1+z_2}
 \end{equation*}

@@

This property does not hold for some other popular losses such as Stein's entropy loss, for example, which prevents their substitution with corresponding validation losses. However, various other losses do conform to this constraints and could be used in this study.
} 
of loss $\loss{C,\Sigma}$.  Thus, estimators resulting in consistently lower validation loss can be inferred to produce estimates that are closer to truth than estimators with higher validation loss.

With empirical data, acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation is used.  In cross-validation, the sample is divided into $K$ subsets of approximately equal size. In all computations in this paper $K=10$ was used.  Then each  subset is used as the validation sample with the other $K-1$ serving as the training dataset. The validation losses from each of such `folds' are averaged to produce \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of  estimator $C$ trained on the remaining $K-1$ subsets. Then cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In the present formulation, $\ell_C$ is known as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier).

>>>>

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser subsets of neurons \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of the functional organization of neural circuits.  However, estimating and interpreting large correlation matrices obtained from finite recordings is challenging.  Estimation can be improved by regularization: biasing of the usual unbiased estimate toward its low-dimensional approximation.  The amount of improvement is greatest when the approximation chosen as the target of regularization parsimoniously captures the dominant dependencies in the data.  Therefore, the selection of the most efficient estimator of neural correlation matrices is an empirical question that informs about the types of dominant dependencies governing the system.

In this study, we sought most statistically efficient estimators of neural correlations in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes of about 100 $\mu$m in diameter (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix would be most efficiently represented by a sparse network of partial correlations between pairs of observed neurons combined with common latent factors representing unobserved inputs and global network fluctuations.  Indeed, in cross-validation tests, the covariance matrix estimator based on such approximations outperformed other regularized estimators. Mirroring previously described patterns of synaptic connectivity, the density of positive partial correlations decreased rapidly with both physical distances and preferred orientation differences whereas negative partial correlations were less selective.  The inferred function connectivity obtained using these methods can generate hypotheses about synaptic architecture that can guide future experimental work.
\section*{Author Summary}
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
Correlations in the activity of neurons have proven useful as descriptors of the functional organization of neural circuits with implications for stimulus coding and circuit architecture.  Estimation of correlation matrices can be improved by imposing some structure on the estimate. Greatest improvements are attained when the imposed structure efficiently captures real dependencies in the data. Using fast 3D two-photon imaging of calcium signals, we recorded the activity of large and dense groups of cells in mouse visual cortex and evaluated the performance of correlation matrix estimators that imposed different kinds of structure. The correlation structure of the estimator that led to greatest improvement comprised a sparse network of partial correlations between pairs of neurons combined with partial correlations with several common latent factors. Both components were necessary for efficient estimation. Since it leads to better estimates, we proposed that this sparse+latent structure provides a better picture of the functional connectivity than that obtained from the usual sample correlation matrix in densely sampled neural recordings. As an application of this approach, we analyzed how the inferred connectivity related to distances between cells and differences in their preferred orientations and found basic agreement with previous studies of synaptic connectivity.

@@

Alternatively, the correlation matrix can be expressed through the partial pairwise correlations. The partial pairwise correlation is the linear correlation remaining after accounting for correlations with all the other neurons. Partial correlations carry particular significance when the relationship between the variables is approximately linear . In this case, small pairwise partial correlations indicate that the two variables are nearly conditionally independent.

Conditional independence between a pair of variables suggests a lack of direct interaction between the underlying processes. Therefore, networks of partial correlations (sometimes called \emph{association networks}) can be used to infer interactions between the components of a complex system.  This approach has been used to uncover the structure of gene interaction networks \cite{Schafer:2005,Peng:2009} and the functional connectivity between brain regions from fMRI signals \cite{Varoquaux:2012,Ryali:2012}. In \emph{elliptical distributions} (those of the form $x \sim \frac 1 Z g\left((x-\mu)^\T\Sigma^{-1}(x-\mu)\right)$, where $g$ is an arbitrary function, $Z$ is the scalar normalizer, $\Sigma$ is the covariance matrix) and the multivariate normal distribution in particular, interactions are completely defined by linear effects and partial correlations express conditional dependencies. A multivariate normal distribution specified by a weighted graph of non-zero partial correlations is known as a Gaussian Graphical Model or Gauss-Markov Random Field \cite{Koller:2009}. With departure from linearity, the correspondence between conditional dependence and partial correlations diminishes and can break down \cite{Loh:2012}. For example, the coupling terms in pairwise Ising models have no direct relationship to the partial correlations \cite{Schneidman:2006,Tkacik:2006}.

\subsection*{Estimation of neural correlation matrices}
Here, we pursued two related aims: (a)~the improved estimation of neural correlation matrices and (b)~discovery of the low-dimensional structure of correlations in recordings of multineuronal activity. An accurate characterization of the patterns of correlations helps in interpreting them, and to better relate them to the architecture and dynamics of the underlying neuronal networks.

The true correlation matrix is the normalized version of the true covariance matrix defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; $x$ is the $p\times 1$ vector of real-valued instantaneous firing rates in bins of duration $\Delta t$; and $\mu$ is the vector of mean firing rates. In the case of noise correlations, $\mu$ depends on the stimulus condition.  The usual estimator of the covariance matrix is the \emph{sample covariance matrix} $C_{\sf 0}$ computed from the empirical sample of observations $x(t),\; t=n(1,\ldots,n)$ as
\begin{equation}
    C_{\sf 0} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ is the number of degrees of freedom per neuron in the sample ($\nu=n-1$ if observations are independent). In this study, we estimate the true mean, $\mu$,  using the sample mean $\bar x$, but seek a better estimate of the covariance matrix than $C_{\sf 0}$.

@@

Estimation of large covariance matrices presents a number of numerical challenges.  The amount of recorded data grows only linearly with population size whereas the number of estimated coefficients increases quadratically.  This mismatch leads to an increase in spurious correlations, overestimation of shared activity (\emph{i.e.}\;overestimation of large eigenvalues) \cite{Ledoit:2004}, and poorly conditioned estimates of the partial correlations \cite{Schafer:2005}.

Estimation can be improved through \emph{regularization}: the deliberate biasing of the estimate toward a low-dimensional approximation \cite{Schafer:2005,Bickel:2006}.  The sample estimate $C_{\sf 0}$ is unbiased ($\E{C_0}=\Sigma$) but, on average, falls far from $\Sigma$ due to its sensitivity to sampling noise.  Low-dimensional estimates are typically less susceptible to sampling noise but can introduce biases.  Regularization strikes a favorable balance between bias and variability, and can produce \emph{some} improvement even with an arbitrary target estimate  Yet, when the target estimate captures important features of the true covariance matrix with few terms, a regularized estimate introduces minimal bias and outperforms other estimators.

\subsection*{Illustration of estimation of a neural correlation matrix}
To illustrate the challenges in estimating the correlation matrix from a finite sample, we consider a regularization scheme based on \emph{covariance selection} (Figure \ref{fig:01}) \cite{Dempster:1972}. Here the estimate is produced by fitting only an optimal subset of the coefficients of the precision matrix while setting the rest to zero. We estimated the covariance matrix of somatic calcium signals of a group of 298 neurons in mouse visual cortex using both the sample covariance matrix (Fig.\;\ref{fig:01}\,A, B) and covariance selection (Fig.\;\ref{fig:01}\,C). Due to the low-pass filtration effect of calcium dye kinetics, correlations in unprocessed calcium signals are higher than typical firing rate correlations on shorter temporal scales. The close similarity of the two estimates of the correlation matrix (Fig.\;\ref{fig:01}\,B and C) belies the utterly different partial correlation structure (Fig.\;\ref{fig:01}\,D, E): The regularized estimate is produced from the same data by setting to zero 31,501 (71.2\%) of the off-diagonal coefficients of the precision matrix and fitting the remaining 12,752 coefficients. Estimation by covariance selection is attractive for several reasons: First, with fewer free parameters it is less susceptible to sampling noise, yielding estimates that are, on average, closer to the truth. Additionally, if we assume that partial correlations do indeed reflect elements of the functional connectivity in the circuit, the second estimate also appears more plausible: in neural circuits, each neuron only interacts with a fraction of the entire population.

@@

\subsection*{Model selection}
Covariance selection is but one of many conceivable low-dimensional approximations of the covariance matrix. Indeed any probabilistic model of the joint population activity regularized toward a small number of parameters automatically defines a low-dimensional approximation of the covariance matrix. Models that accurately reflect the dominant regularities in the activity of the circuit, its so-called `functional connectivity', tend to produce estimates that are nearest to the true value of the covariance matrix and are said to be more \emph{efficient} or to \emph{dominate} other estimators. An estimator's superior efficiency can be ascertained by cross-validation.  Because different types of neural circuits have different types of regularities in their activity, different probabilistic models and covariance matrix estimators may dominate in each case. The estimator that is shown to dominate all others for a specific type of neural circuit may provide the best descriptor of the low-dimensional representation of the correlation structure of the system of interest. Its structure may then be analyzed and related to the circuit's anatomical organization.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Estimators whose low-dimensional regularization targets can represent the structure of the true covariance matrix outperform other estimators.}
        {\bf Row 1.} Graphical representations of four types of low-dimensional structures of interactions between observed neurons (green spheres) and latent units (light-shaded spheres).
        In the `independent model' ({\bf A}), observed neurons exert no linear effects on one another neither directly nor through interactions with common latent units. 
        In `latent factors' ({\bf B}), the correlated activity of observed cells is driven by several latent units. 
        In `sparse interactions' ({\bf C}), the correlation matrix is defined by a set of linear interactions between observed neurons. 
        In `sparse+latent' ({\bf D}), correlations arise through direct linear interactions in a sparse subset of observed neurons and through interactions with common latent units. 
        {\bf Row 2.} Examples of $50\times 50$ correlation matrices corresponding to each type of low-dimensional structure. 
        The factor model ({\bf B}) has three latent units. 
        The partial correlation matrix of the sparse model ({\bf C}) is 73\% sparse.
        The `sparse+latent' matrix has one latent unit and its direct interactions are 78\% sparse.
        {\bf Row 3.} Examples of sample correlation matrices calculated from samples of 1000 observations taken from simulated random processes with corresponding correlation matrices from row 2.
        {\bf Row 4.} Mean \emph{excess loss} (Eq.~\ref{eq:excess-loss}) attained by each of the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.
        {\bf Row 5.} Mean \emph{cross-validation loss} (Eq.~\ref{eq:rel-cv-loss}) of covariance estimators with respect to the matching estimator. The values are relative to the validation loss of the estimator that matches the low-dimensional structure of the true covariance matrix. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

For the first regularized estimator $C_{\sf diag}$, the target estimate is the diagonal matrix $D$ whose diagonal contains  variance estimates. Regularization is achieved by linear \emph{shrinkage} of the unbiased estimate $C_{\sf 0}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}
C_{\sf diag} = (1-\lambda) C_{\sf 0} + \lambda D
\end{equation}
The diagonal target has structure corresponding to the absence of linear associations between the activity of observed neurons (Fig.~\ref{fig:02}\,A).  If this structure accurately describes recorded data, then strong shrinkage toward $D$ adds little bias while strongly reducing the variability of the estimate. Shrinkage allows for partial commitment to the low-dimensional representation.

The target of the second regularized estimator $C_{\sf factor}$, is the factor model $F =  L L^\T + \Psi$ with $d$ factors. Therefore $L$ is the $p\times d$ matrix of \emph{factor loadings} and the diagonal matrix $\Psi$ contains the independent variances of each neuron. The estimator, 
\begin{equation}
C_{\sf factor} = (1-\lambda) C_{\sf 0} + \lambda F,
\end{equation}
has two hyperparameters: the number of factors $d$ and shrinkage intensity $\lambda$. The target estimate $F$ has the structure that arises when correlated fluctuations in population activity are driven by a small number of latent factors that affect many cells while direct interactions between cells are insignificant (Fig.~\ref{fig:02}\,B).

The third estimator $C_{\sf sparse}$ is based on the assumption that all correlations result from direct linear associations between a sparse set of pairs of observed cells (Fig.~\ref{fig:02}\,C).  This assumption is enforced by setting to zero an optimal subset of the off-diagonal elements of the precision matrix.  Therefore, $C_{\sf sparse}$ is biased towards correlation structures that arise in networks of linearly interacting pairs of neurons (Fig.~\ref{fig:02}\,C). The estimator takes the form
\begin{equation}
C_{\sf sparse} = S^{-1},
\end{equation}
where $S$ is a sparse matrix with a large fraction of zeros off the diagonal. The estimate has one hyperparameter that determines the sparsity (fraction of off-diagonal zeros) of $S$.

The fourth estimator $C_{\sf sparse+latent}$, combines the assumptions of the previous two.  The underlying structure arises from sparse interactions between the recorded neurons and latent factors driving their activity (Fig.~\ref{fig:02}\,D). The estimator has the form \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}
C_{\sf sparse+latent} = (S - LL^\T)^{-1}
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a $d\times p$ matrix of factor loadings. The estimator has two hyperparameters: the number of latent units $d$ and the sparsity of $S$.

\subsection*{Simulation}
To illustrate the performance of the four regularized estimators, we constructed four model populations of $p=50$~neurons each with each matching the correlation structures of a different target estimate (Fig.~\ref{fig:02} Row 2).  Sample correlation matrices calculated from samples of size $n=1000$ contain substantial sampling noise (Fig.~\ref{fig:02) Row 3) that makes the correlation structure less obvious.

To evaluate the performance of a covariance matrix estimator $C$, we define a real-valued \emph{loss function} $\loss{C,\Sigma}$ that attains a minimum when $C=\Sigma$.  The loss function quantifies the error of the estimate, \emph{i.e.}~the deviation of the estimate $C$ from truth $\Sigma$.

@@

The choice of loss function is motivated by mathematical convenience. We expect that the main conclusions of our study will not change qualitatively with other well behaved loss functions, such as the Frobenius norm of the difference $\|C-\Sigma\|_F$ \cite{Ledoit:2004,Schafer:2005}, Stein's entropy loss, or quadratic loss \cite{James:1961,Fan:2008}.

We define the \emph{excess loss} as
\begin{equation}\label{eq:excess-loss}
    \eloss{C,\Sigma} = \loss{C,\Sigma}-\loss{\Sigma,\Sigma},
\end{equation}
which assumes zero at its minimum.

In Figure \ref{fig:02}, row 4 shows the mean excess losses and their standard errors calculated from 30 samples of sizes n=250, 500, 1000, 2000, and 4000 for each estimator and each ground truth. For each estimator, hyperparameters were estimated by nested cross-validation (See Methods for more details.)

As expected, estimators with whose structure matched that of the true model typically outperformed the other estimators. There were two exceptions: First, for small sample sizes, there was insufficient data to reveal the true correlation structure. In this case estimates with simpler targets often outperformed the estimator based on the correct model. Second, the sparse+latent estimator often performed equally well to the sparse estimator even when the data did not have any latent units.  In these cases, it inferred the right number of latent units: zero. In larger models, the falsely identified latent factors are more common.

When the ground truth, $\Sigma,$ is not accessible, loss can be estimated solely from the data through \emph{validation}.  In validation, an additional independent \emph{testing sample} is used to compute a sample covariance estimate $C_{\sf 0}^\prime$.  Then \emph{validation loss} $\loss{C,C_{\sf 0}^\prime}$ can be used as an unbiased estimate\footnote
{
    Validation loss $\loss{C,C_{\sf 0}^\prime}$ is an unbiased estimate of loss $\loss{C,\Sigma}$ when $\loss{\cdot,\cdot}$ is \emph{additive} in its second argument so that 
 \begin{equation*}\label{eq:additivity}
 \loss{C,z_1} + \loss{C,z_2} \equiv \loss{C,z_1+z_2}
 \end{equation*}

@@

This property does not hold for some other popular loss functions such as Stein's entropy loss, for example, which prevents their substitution with corresponding validation losses. However, various other loss functions do have this property and could have been used.
} 
of loss $\loss{C,\Sigma}$.  Thus, estimators resulting in consistently lower validation loss can be inferred to produce estimates that are closer to truth than estimators with higher validation loss.

With empirical data, acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation is used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each  subset is used as the validation sample with the other $K-1$ serving as  a training dataset. The validation losses from each of such `folds' are averaged to produce \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of  estimator $C$ trained on the remaining $K-1$ subsets. Then cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In the present formulation, $\ell_C$ is known as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier).
